---
title: Advanced Network Analysis
subtitle: Random Graphs
author: "Olga Chyzh [www.olgachyzh.com]"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
exclude: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', message = FALSE, warning = FALSE)
```

## Why ERGMs?

The fundamental problem of network analysis:

- Is the network we observe an Erdos-Renyi network?

    - In an Erdos-Renyi network, the probability of each edge is independent of that of other edges.
    
- If not, what are the endogeneous network features that define our network?
---
## Example: Friendship Netwok

```{r, echo=TRUE, message=FALSE}
library(igraph)
library(sna)
data(coleman) #Use friendship data
coleman<-coleman[1,,]
#convert to an -igraph- object, we'll treat it as a directed graph for now:
coleman<-graph_from_adjacency_matrix(coleman, mode="directed", diag=FALSE) 
edge_density(coleman)
reciprocity(coleman) #Note that -igraph- default is an undirected graph

```

---
## Friendship Data
```{r, eval=TRUE, echo=FALSE,out.width = '60%'}
plot(coleman, vertex.label=NA, vertex.color="blue",  vertex.size = 5,edge.arrow.size = 0.5)
```

- What are some endogenous network features of this network?
---
## Erdos-Renyi Networks

- In order to answer whether an observed network is different from a random network, it would help to know what a random network would look like.

- In a random network, all edges have the same probability of realization, $p$. Moreover, the probability of realization of edge $i$, $p_i$,  does not depend on $p_j$.

---

## Simulate A Random Graph to Compare to the Coleman Data

Need: 

- number of nodes $N$ in the friendship network
- the probability that each two nodes are connected $p$

```{r}
vcount(coleman) #returns the number of vertices
gsize(coleman) #returns the number of edges

#or
summary(coleman)
```



---
## Simulate A Random Graph to Compare to the Coleman Data

- Let $p$ denote the probability that any two vertices are connected by an edge.  Then, we can calculate the value of $p$ in the friendship network as the number of observated edges over the number of all possible edges.

  + With 73 vertices, we have $73*72/2=2628$ possible undirected edges or $73*72=5256$ directed edges, as each vertex can connect to each other vertex, but there are no self-loops. (Since the friendship network is directed, let's focus on simulating a directed network.)

  + Then $p=243/5256=0.046$. Does this value seem high or low? Note that $p$ is the clustering coefficient of a random network with a given number of nodes and edges.

---

## Simulate A Random Graph (Continued)

```{r, echo=TRUE, eval=TRUE, message=FALSE}
set.seed(45765) #since a simulation involves randomness, set the seed for 
#reproducibility.
#Step i--start with a matrix of 73 unconnected nodes.
N=73 #Set the number of nodes:
rnet<-matrix(0, nrow=N,ncol=N) 

#Step ii:
p<-243/5256
for (i in 1:N) {
  for (j in 1:N){
    if (i!=j) {
    rnet[i,j]=as.numeric(runif(1)<p)
    }}}

```
---





## Check Our Work

```{r, echo=TRUE, eval=TRUE, message=FALSE}
summary(g<-graph_from_adjacency_matrix(rnet, mode="directed", weighted=NULL))

edge_density(coleman)
reciprocity(coleman) #Note that -igraph- default is an undirected graph

edge_density(g)
reciprocity(g) #Note that -igraph- default is an undirected graph


```

---
## Visualize

```{r, eval=TRUE, echo=FALSE,out.width = '80%'}
par(mfrow = c(1, 2))
plot(coleman, vertex.label=NA, vertex.color="blue",  vertex.size = 5,edge.arrow.size = 0.05)
title("Actual")

plot(g, vertex.label=NA, vertex.color="blue",  vertex.size = 5,edge.arrow.size = 0.05)
title("Random")
```
---
## Summarize the Observed and Simulated Networks

```{r, eval=TRUE, echo=FALSE}
library(statnet)
data(coleman)
coleman<- as.network.matrix(coleman[1,,],	matrix.type='adjacency', directed=TRUE)
rnet<- as.network.matrix(rnet, matrix.type='adjacency',	directed=TRUE)
summary(coleman~ edges+idegree(6)+ triangles+ mutual+ostar(2)+istar(2))
summary(rnet~edges+idegree(6)+ triangles+ mutual+ostar(2)+istar(2))

```

---
## Your Turn

1. Simulate a random network that we could compare to the Sampson data.

2. Plot the two side-by-side

3. Use summary to further explore the differences between the two.

4. Based on this analysis, what model specification would you propose?

