---
title: "Foundations"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    dev = "png", dpi = 150,
    cache = FALSE,
    echo = TRUE
)
```

##  Package Overview 

You supply data and `netify` transforms it into easy to work with network data. The goal of `netify` is to provide R functions that simplify and facilitate common tasks related to network creation, summary, visualization, and modeling. Although our package was built with social scientists (especially peace science scholars) in mind, anyone can use it!

This vignette provides a high level overview of our package from start to finish. The best use of this vignette is to introduce the core components of the package to larger audiences. This overview covers our primary functions with data examples and minimal writing. 

`netify` goals: 

  (1) Create: Netify your data! We Make it easy for users to create networks from raw data in $R$ as well as add additional features, such as nodal and dyadic variables, to the network object.
  (2) Explore: Explore characteristics of the network you created, like summary statistics at both the network and actor levels. Visualize your network.
  (3) Advance:  Advance your network analysis to the next stage by preparing it for use in other network packages and modeling approaches. 
  
`netify` provides a suite of primary functions to help achieve these goals:

| Create💡  | Explore 🔎 | Advance ️🚀|
|---------|---------|---------|
| `netify()`  | `peek()`  | `netify_to_amen()`  |
| `add_node_vars()` | `summary_actor()`   | `netify_to_igraph()`  |
| `add_dyad_vars()`  | `summary()`  | `netify_to_statnet()`  |
| `subset_netify()`  | `plot_actor_stats()`  |   | 
|              |   `plot_graph_stats()`|        |
|              |   `plot()`|        |

`netify` begins with the user's data input. Data types: our core function, `netify`  handles up to 7 different data inputs, including data frames and edgelists, see documentation for more detail.

- The package can also create different types of networks including:

  - cross sectional networks 
  - longitudinal (with static and varying actor composition)
  - bipartite networks
  - multilayer

- As well as create networks with different edge types:
  - weighted 
  - binary 
  - symmetric or non-symmetric 

## Step 1: Create 💡

Begin by loading packages and supplying the data. We will use the `peacesciencer` package to grab some familiar data.


```{r, eval=TRUE, results='hide', warning=FALSE, message=FALSE}
# load packages
library(netify)

# install extra packages for this vignette
if (!"tidyverse" %in% rownames(installed.packages())) {
    install.packages("tidyverse", repos = "https://cloud.r-project.org")
}
if (!"peacesciencer" %in% rownames(installed.packages())) {
    install.packages("peacesciencer", repos = "https://cloud.r-project.org")
}
# load necessary packages for this vignette
library(peacesciencer)
library(tidyverse)

# organize external data for peacesciencer
peacesciencer::download_extdata()

# create dyadic data set over time using peacesciencer
cow_dyads <- create_dyadyears(
    subset_years = c(1995:2014)
) |>
    # add mids
    add_cow_mids() |>
    # add capital distance
    add_capital_distance() |>
    # add democracy
    add_democracy() |>
    # add gdp
    add_sdp_gdp()
```

Next, create a `netlet` object from the above COW data frame using our package's core function `netify.` There are a number of useful parameters, but the most important ones to highlight are:

- `input` is, in this use case, a dyadic data.frame that should have at the following variables used to specify actors:

  - `actor1`: indicates actor 1 variable in the data
  - `actor2`: indicates actor 2 variable in the data
  
- `netify_type` is a type of netlet object (`cross-sec`, `longit_list`, or `longit_array`).

```{r, eval=TRUE}
mid_long_network <- netify(
    input = cow_dyads,
    actor1 = "ccode1", actor2 = "ccode2", time = "year",
    weight = "cowmidonset",
    actor_time_uniform = FALSE,
    sum_dyads = FALSE, symmetric = TRUE,
    diag_to_NA = TRUE, missing_to_zero = TRUE,
    nodal_vars = c("v2x_polyarchy1", "v2x_polyarchy2"),
    dyad_vars = c("capdist"),
    dyad_vars_symmetric = c(TRUE)
)
```

**Congratulations** you have created a network object! 🎉

We can also add nodal and dyadic data after we've created the network via the `add_node_vars()` and `add_dyad_vars()` functions. 

Let's assume that we had information about each actor in the network that we'd like to add as a nodal variable after we already made the network object. This could be from our original data set or elsewhere. For example, lets add a logged variable measuring gdp for each node in the network over time:

```{r, eval=TRUE}
# create a vector of nodal data
node_data <- unique(cow_dyads[, c("ccode1", "year", "wbgdppc2011est2")])
node_data$wbgdppc2011est2_log <- log(node_data$wbgdppc2011est2)

# add nodal variable to netlet object
mid_long_network <- add_node_vars(
    netlet = mid_long_network,
    node_data = node_data,
    actor = "ccode1",
    time = "year"
)

# create another dyadic var in cow
cow_dyads$log_capdist <- log(cow_dyads$capdist + 1)

# now lets add this to the netlet
mid_long_network <- add_dyad_vars(
    netlet = mid_long_network,
    dyad_data = cow_dyads,
    actor1 = "ccode1",
    actor2 = "ccode2",
    time = "year",
    dyad_vars = "log_capdist",
    dyad_vars_symmetric = TRUE
)
```

## Step 2: Explore and Summarize 🔎

We made a network, so let's look at it. First, we might want to take a `peek` at the network object to see if the matrix looks the way we'd expect it to look. This function lets you glance at a specific slice of the network if it is longitudinal or the entire network if it is cross-sectional. (To actually subset the netlet object and make a new object use netify's `subset` function.)

```{r}
peek(mid_long_network,
    from = 5, to = 5,
    time = c("2009", "2010")
)
```

Next, let's examine a few basic summary statistics about the network using our`summary()` function.

```{r}
# create data.frame that provides network-level summary stats
# for each year of the network
mid_long_summary <- summary(mid_long_network)
```

We can also make a quick visualization of network statistics over time using the summary statistics data frame.

```{r, fig.width=6, fig.height=6}
plot_graph_stats(mid_long_summary)
```

These graph statistics are useful for understanding changes overtime at the network level, however, we might also want to look at actor-level statistics overtime. Our built-in function, `summary_actor` will calculate in-degree, out-degree, average degree, and eigenvector centrality for each actor in each time period. The `across_actors` function allows users to toggle whether they want a summary of a given statistic across all actors (shown in a density plot) or for specific actors:

```{r}
# every year & every actor
summary_actor_mids <- summary_actor(mid_long_network)
head(summary_actor_mids)
```

We can look at the distribution of the statistic for all actors over time:

```{r, fig.width=6, fig.height=6, warning = FALSE, message = FALSE}
# density plot across all actors
# for each stat
plot_actor_stats(
    summary_actor_mids,
    across_actor = TRUE,
)
```

Or we might like to select a specific statistic to focus on across actors over time:

```{r, fig.width=6, fig.height=6}
# focus on closeness
plot_actor_stats(
    summary_actor_mids,
    across_actor = TRUE,
    specific_stats = "closeness"
)
```

Say that we wanted to showcase actor-specific statistics over time. We can use the plot_actor_stats function for this as well, though it's **highly recommended** to subset to a few actors for a more legible plot.

```{r, fig.width=6, fig.height=6}
# top 5 GDP countries (USA, China, Japan, Germany, India)
top_5 <- c("2", "710", "740", "255", "750")

#
plot_actor_stats(
    summary_actor_mids,
    across_actor = FALSE,
    specific_actors = top_5
)
```

We can also zoom into a specific time slice of the network:

```{r, fig.width=6, fig.height=6}
summary_df_static <- summary_actor_mids[summary_actor_mids$time == 2011, ]

plot_actor_stats(
    summary_df_static,
    across_actor = FALSE,
    specific_actors = top_5
)
```

Instead of looking at summary statistics, we also might want to simply visualize the entire network. We can do this by plotting the netify object. (Isolates removed by default and seed set to 6886 for node layout).

Our goal was to make this super easy for plotting time-varying attributes or actors. 

```{r, fig.width=6, fig.height=6}
# default plot
plot.netify(mid_long_network,
    static_actor_positions = TRUE,
    remove_isolates = FALSE
)

# a little cleaner
plot(
    mid_long_network,
    edge_color = "grey",
    node_size = 2
)
```

We can also use `netify` functions to explore actor level summary statistics in the network graph. 

```{r}
# add actor variables from summary_actor_mids
mid_long_network <- add_node_vars(
    mid_long_network,
    summary_actor_mids,
    actor = "actor", time = "time",
    node_vars = c("degree", "prop_ties", "eigen_vector"),
)
```

We can quickly inspect the object with the `print` function.


```{r}
# print netlet obj to make sure they got added to nodal features
print(mid_long_network)
```


As well as look at the attributes of the data and specify showing the nodal data information.

```{r}
# if you're curious as to where they live
head(
    attr(
        mid_long_network,
        "nodal_data"
    )
)

# i.e.,
head(attributes(mid_long_network)$nodal_data)
```

And now return to our network graph by highlighting specific nodal attributes:

```{r, fig.width=6, fig.height=6}
# vary node size by degree
plot(
    mid_long_network,
    edge_color = "grey",
    point_size_var = "degree"
)

# vary node color by polyarchy
plot(
    mid_long_network,
    edge_color = "grey",
    node_size_by = "degree",
    node_color_by = "v2x_polyarchy1",
    node_color_label = "Polyarchy",
    node_size_label = "Degree"
) +
    scale_color_gradient2()
```

We might also prefer to add labels, but only a select few:

```{r, fig.width=6, fig.height=6}
library(countrycode)
cowns <- countrycode(
    c(
        "United States", "China", "Russia",
        "France", "Germany", "United Kingdom"
    ),
    "country.name", "cown"
)
cabbs <- countrycode(cowns, "cown", "iso3c")

plot(
    mid_long_network,
    edge_color = "grey",
    node_size_by = "degree",
    node_color_by = "v2x_polyarchy1",
    node_color_label = "Polyarchy",
    node_size_label = "Degree",
    select_text = cowns,
    select_text_display = cabbs,
    text_size = 3
) +
    scale_color_gradient2()

# or we can go with labels only
# and remove points
plot(
    mid_long_network,
    edge_color = "grey",
    add_points = FALSE,
    add_label = TRUE,
    label_size_by = "degree",
    label_color = "white",
    label_fill_by = "v2x_polyarchy1",
    label_fill_label = "Polyarchy",
    label_size_label = "Degree"
) + guides(size = "none")
```

## Step 3: Advance 🚀

Once we have created and explored our network object, we might want to continue analyzing the data using different modeling approaches. `netify` makes this simple even though those models often aren't! And for the sake of convergence lets go with cross-sectional networks. 

First, prep the data:

```{r, warning=FALSE, message = FALSE}
# prep data
cow_cross <- cow_dyads |>
    group_by(ccode1, ccode2) |>
    summarize(
        cowmidonset = ifelse(any(cowmidonset > 0), 1, 0),
        capdist = mean(capdist),
        polity21 = mean(polity21, na.rm = TRUE),
        polity22 = mean(polity22, na.rm = TRUE),
        wbgdp2011est1 = mean(wbgdp2011est1, na.rm = TRUE),
        wbgdp2011est2 = mean(wbgdp2011est2, na.rm = TRUE),
        wbpopest1 = mean(wbpopest1, na.rm = TRUE),
        wbpopest2 = mean(wbpopest2, na.rm = TRUE)
    ) |>
    ungroup() |>
    mutate(
        capdist = log(capdist + 1)
    )

# subset set to actors with 10mil pop
actor_to_keep <- cow_cross |>
    select(ccode1, wbpopest1) |>
    filter(wbpopest1 > log(10000000)) |>
    distinct(ccode1)

# filter cow_cross by actor_to_keep
cow_cross <- cow_cross |>
    filter(ccode1 %in% actor_to_keep$ccode1) |>
    filter(ccode2 %in% actor_to_keep$ccode1)

# create netlet
mid_cross_network <- netify(
    cow_cross,
    actor1 = "ccode1", actor2 = "ccode2",
    weight = "cowmidonset",
    sum_dyads = FALSE, symmetric = TRUE,
    diag_to_NA = TRUE, missing_to_zero = FALSE,
    nodal_vars = c(
        "polity21", "polity22", "wbgdp2011est1",
        "wbgdp2011est2", "wbpopest1", "wbpopest2"
    ),
    dyad_vars = c("capdist"),
    dyad_vars_symmetric = c(TRUE)
)
```

Next, let's take a look at passing our netify object to the `amen` function:

```{r}
# install (if necessary) and load amen
if (!"amen" %in% rownames(installed.packages())) {
    install.packages("amen", repos = "https://cloud.r-project.org")
}
library(amen)

# prep for amen
mid_cross_amen <- netify_to_amen(mid_cross_network)

# we got all the elements we need for amen! woohoO!
str(mid_cross_amen)

# plug and run
mid_amen_mod <- ame(
    Y = mid_cross_amen$Y,
    Xdyad = mid_cross_amen$Xdyad,
    Xrow = mid_cross_amen$Xrow,
    family = "bin",
    R = 0,
    symmetric = TRUE,
    seed = 6886,
    nscan = 50,
    burn = 10,
    odens = 1,
    plot = FALSE,
    print = FALSE
)
```

We can apply the same process to ERGMs:

```{r}
# install (if necessary) and load ergm
if (!"ergm" %in% rownames(installed.packages())) {
    install.packages("ergm", repos = "https://cloud.r-project.org")
}
library(ergm)

# called netify_to_statnet because it's a reference
# to the network library, which is what ergm uses
mid_cross_ergm <- netify_to_statnet(mid_cross_network)

# attributes should all be loaded into the
# appropriate slot
# notice edge attribtues get a _e suffix added
mid_cross_ergm

# set NA values to 0 for the three nodecov variables
# this is only for demonstration purposes in the vignette/example
# in any real analysis, carefully consider how to handle missing data
set.vertex.attribute(
    mid_cross_ergm, "polity21", 
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "polity21")), 
        0, get.vertex.attribute(mid_cross_ergm, "polity21")))

set.vertex.attribute(
    mid_cross_ergm, "wbgdp2011est2", 
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "wbgdp2011est2")), 
        0, get.vertex.attribute(mid_cross_ergm, "wbgdp2011est2")) )

set.vertex.attribute(
    mid_cross_ergm, "wbpopest2", 
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "wbpopest2")), 
        0,  get.vertex.attribute(mid_cross_ergm, "wbpopest2")) )

# plug and run
# Fit the ERGM model (well not a real ergm)
ergm_model <- ergm(
    formula = mid_cross_ergm ~
        edges +
        nodecov("polity21") +
        nodecov("wbgdp2011est2") +
        nodecov("wbpopest2")
)
```



## References

- Csárdi G, Nepusz T, Traag V, Horvát S, Zanini F, Noom D, Müller K (2024). igraph: Network Analysis and Visualization in R. doi:10.5281/zenodo.7682609, R package version 2.0.3, https://CRAN.R-project.org/package=igraph.

- Davies, Shawn, Therese Pettersson & Magnus Öberg (2023). Organized violence 1989-2022 and the return of conflicts between states?. Journal of Peace Research 60(4).

- Handcock M, Hunter D, Butts C, Goodreau S, Krivitsky P, Morris M (2018). ergm: Fit, Simulate and Diagnose Exponential-Family Models for Networks. The Statnet Project (http://www.statnet.org). R package version 3.9.4, https://CRAN.R-project.org/package=ergm.

- Hoff, Peter D. "Dyadic data analysis with amen." arXiv preprint arXiv:1506.08237 (2015).

- Högbladh Stina, 2023, “UCDP GED Codebook version 23.1”, Department of Peace and Conflict Research, Uppsala University

- Miller S (2022). “peacesciencer: An R Package for Quantitative Peace Science Research.” Conflict Management and Peace Science, 39(6), 755–779. doi: 10.1177/07388942221077926.

- Statnet Development Team (Pavel N. Krivitsky, Mark S. Handcock, David R. Hunter, Carter T. Butts, Chad Klumb, Steven M. Goodreau, and Martina Morris) (2003-2023). statnet: Software tools for the Statistical Modeling of Network Data. URL http://statnet.org

- Sundberg, Ralph and Erik Melander (2013) Introducing the UCDP Georeferenced Event Dataset. Journal of Peace Research 50(4).
