---
title: 'Network Analysis II - Homework #4'
author: "Your Name"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Make sure to replace "Your Name" in the YAML (top of this script) with your name.

## International Alliance network

2. You will be using the data `ally_data` from the `networkdata` package. Run the code in the chunk below to load the data. Remember to update the package to the most recent version.  

```{r, message=F}
library(devtools)
install_github("ochyzh/networkdata")
library(networkdata)
data(ally_data)


```

3. Check the class and the dimensions of each object that makes up the data. Look at the first 10 rows and columns of `W` and the first 10 rows of `ally_data`. Type your code in the chunk below. 

```{r}
#Type code here.

```


4. In class, we wrote the code to estimate an lsgm with 1 edge-level covariate X using a pseudo-likelihood.  Edit the chunk below so that the resulting pseudo-likelihood can include 2 edge-level covariates X1 and X2 (i.e., the goal is to run an lsgm with `tot_trade` and `mil_ratio` as exogenous covariates). 


```{r}
#Type code here:
#Likelihood
loglik<-function(par,X,W,Y){
b0<-par[1]
b1<-par[2]
eta<-par[3]
xbeta<-b0+b1*X
kappa<-exp(xbeta)/(1+exp(xbeta)) #logit of Xb
A_i=log(kappa/(1-kappa))+eta*W%*%(Y-kappa) #Eqn 2
p_i<- exp(A_i)/(1+exp(A_i)) #Eqn 1, also Eqn 4
PL<-Y*log(p_i)+(1-Y)*log(1-p_i) #Eqn 3
ell <- -sum(PL)
#cat("ell",ell, fill=TRUE)
return(ell)
}


```


5. Estimate an lsgm with `defense` as the dependent variable, with `tot_trade` and `mil_ratio` as the exogenous edge-level covariates, and `W` as the matrix that measure connectivities among defense edges. 

```{r}
#Type code below:


```

6. Looking just at the direction of the coefficients, do you find the same results as Chyzh and Kaiser (2019)?

7. Ungraded *Bonus* question: change the code from class to estimate the standard errors for your model.

```{r}
#Type code here:


```


8. Knit the RMarkdown file (press the "Knit" button above) and read through the corresponding html file.  

Note: 

This homework is supposed to be mainly a code reading exercise with a little bit of coding. You simply have to figure out which pieces you need to copy and adjust. 


For the submission: submit your solution in an R Markdown file and (just for insurance) submit the corresponding html file with it. 
